### 2021-02-09 final_report/include/TT.tex
reviewed TT chapter

While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate. In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we'll see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}. \smallskip

@@

% intro to GF
Grammatical Framework (GF), being a grammar formalism and programming language designed with this parallel in mind, goes a step further in the direction of synchronous CFGs and TAGs by introducing a clear distinction between the \textit{abstract syntax}, whose aim is to capture the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, consisting in their linearization rules \cite{gf2004} \cite{gfbook}. 
This makes it possible to deal with multiple languages by writing only one grammar, whose components are an abstract syntax and several concrete syntaxes, providing a solid basis for a system able to translate between any pair of for which a concrete syntax is available, in any direction. \smallskip

% GF as an interlingua-based MT system
With respects to this, a GF can be also seen as an \textit{interlingual} MT system where the intermediate representation or \textit{interlingua} is an AST. Interlingua-based systems have the advantage, in terms of efficiency, of making it unnecessary to build $n (n - 1)$ translation functions to cover all possible pairs of $n$ languages: having $2n$ is in general enough. In GF in particular, using a multilingual grammar makes it so that translation can be performed, for each pair of languages, in both directions, so that $n$ translation functions (or rahter $n$ concrete syntaxes and an abstract syntax) are already sufficient.\smallskip

@@

\subsection{Dependency grammars} \label{depg}
% DGs as opposed to PSGs
A class of grammar formalisms alternative to phrase structure, first proposed in \cite{dg}, is that of \textit{dependency grammars}, the main difference between the two being the relation they are based on. 
As opposed to constituency, \textit{dependency} is a word-to-word correspondence, meaning that words are simply put in relation with each other via directed links, called in fact \textit{dependencies}. Each dependency is, then, composed of two words: a \textit{head} and a \textit{dependent} that refers to it. For instance, if we try to look at syntactic dependencies in the sentence ``\textit{grammars rule}'', we could identify the verb ``\textit{rule}'' as the head and the noun ``\textit{children}'', its subject, as its dependent. \smallskip

% pros
Intuitively, this makes it so that dependency trees are simpler, which makes them an easier target for the frontend, possibly ML-based, of a MT system such as the one outlined in the above. 
Existing \textit{dependency parsers}, such as UDPipe \cite{udpipe1} and the Standford parser \cite{standford}, are often - but not always \cite{rasp} - neural pipelines trained on dependency treebanks. As a result, they are significantly more robust than their phrase structure counterparts.

% cons
On the other hand, there is currently no effective way to use these trees as the starting point for NLG: dependency grammars are an effective way to describe language, but are not \textit{generative}, unlike their phrase-structure counterparts\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}. \smallskip

One idea is, then, to develop a hybrid system where the frontend is a dependency parser and the backend a grammaticality-preserving GF-based target language generation module. 
The connecting link between these two seemingly incompatible stages of our MT pipeline can be, as we will elaborate on in Section \ref{ourapproach}, a CA component able to identify matching dependency trees and generate the corresponding GF concrete and abstract syntax functions: concepts.

@@

In the following, we give a quick overview of the standard format UD uses to store dependency trees and of the aspects of the annotation scheme that are most relevant to the task at hand. Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}. The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{\url{universaldependencies.org}}.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}{.55\textwidth}
      \centering
      \footnotesize
        \begin{verbatim}
1	this	this	PRON	_	_	5	nsubj	_	_
2	is	be	AUX	_	_	5	cop	_	_
3	a	a	DET	_	_	5	det	_	_
4	dependency	dependency	NOUN	_	_	5	compound	_	_
5	tree	tree	NOUN	_	_	0	root	_	_
        \end{verbatim}
    \end{subfigure}%
    \begin{subfigure}{.45\textwidth}
        \centering
        \footnotesize
        \setlength{\unitlength}{0.21mm}
        \begin{picture}(371.0,130.0)
            \put(0.0,0.0){this}
            \put(46.0,0.0){is}
            \put(83.0,0.0){a}
            \put(120.0,0.0){dependency}
            \put(220.0,0.0){tree}
            \put(0.0,15.0){{\tiny PRON}}
            \put(46.0,15.0){{\tiny AUX}}
            \put(83.0,15.0){{\tiny DET}}
            \put(120.0,15.0){{\tiny NOUN}}
            \put(220.0,15.0){{\tiny NOUN}}
            \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
            \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
            \put(105.0,99.66666666666667){{\tiny nsubj}}
            \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
            \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
            \put(128.0,83.0){{\tiny cop}}
            \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
            \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
            \put(146.5,66.33333333333334){{\tiny det}}
            \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
            \put(131.5,35.0){\vector(0,-1){5.0}}
            \put(165.0,49.66666666666667){{\tiny compound}}
            \put(235.0,130.0){\vector(0,-1){100.0}}
            \put(240.0,120.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A dependency tree in CoNNL-U format alongside its graphical representation]{A dependency tree in CoNNL-U format alongside its graphical representation. Optional fields in the CoNLL-U file are left blank.}
    \label{deptree}
\end{figure}

\paragraph{Universal POS tags} \label{upos}
Universal POS tags mark the core part-of-speech categories, such as nouns, verbs, pronouns and determiners. An important - as we will see in Section \ref{criteria} - distinction we can make based on POS tags is that between \textit{content} or \textit{open class} words, i.e. words that contribute to the meaning of the sentences they appear in (nouns, lexical verbs, adjectives, adverbs and interjections), and \textit{function} or \textit{closed class} words, like pronouns and determiners, so called because they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}. \smallskip

@@

 If a sentence contains is no verbs at all, there is no fixed rule excepts that in order to avoid unnecessary discrepancies between languages, the root should be a content word. \smallskip

\begin{figure}[H]

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
\begin{picture}(259.0,90.0)
  \put(0.0,0.0){Katia}
  \put(55.0,0.0){reads}
  \put(110.0,0.0){a}
  \put(147.0,0.0){magazine}
  \put(0.0,15.0){{\tiny PROPN}}
  \put(55.0,15.0){{\tiny VERB}}
  \put(110.0,15.0){{\tiny DET}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(37.5,30.0){\oval(49.54545454545455,33.333333333333336)[t]}
  \put(12.727272727272727,35.0){\vector(0,-1){5.0}}
  \put(22.5,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny det}}
  \put(121.0,30.0){\oval(88.73913043478261,66.66666666666667)[t]}
  \put(165.3695652173913,35.0){\vector(0,-1){5.0}}
  \put(106.0,66.33333333333334){{\tiny obj}}
  \put(70.0,90.0){\vector(0,-1){60.0}}
  \put(75.0,80.0){{\tiny root}}
\end{picture}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
        \begin{picture}(344.0,90.0)
            \put(0.0,0.0){A}
            \put(37.0,0.0){magazine}
            \put(119.0,0.0){is}
            \put(156.0,0.0){read}
            \put(202.0,0.0){by}
            \put(239.0,0.0){Katia}
            \put(0.0,15.0){{\tiny DET}}
            \put(37.0,15.0){{\tiny NOUN}}
            \put(119.0,15.0){{\tiny AUX}}
            \put(156.0,15.0){{\tiny VERB}}
            \put(202.0,15.0){{\tiny ADP}}
            \put(239.0,15.0){{\tiny PROPN}}
            \put(28.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(14.054054054054054,35.0){\vector(0,-1){5.0}}
            \put(13.5,49.66666666666667){{\tiny det}}
            \put(106.5,30.0){\oval(116.47899159663865,66.66666666666667)[t]}
            \put(48.260504201680675,35.0){\vector(0,-1){5.0}}
            \put(91.5,66.33333333333334){{\tiny nsubj:pass}}
            \put(147.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(133.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(132.5,49.66666666666667){{\tiny aux:pass}}
            \put(230.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(216.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(215.5,49.66666666666667){{\tiny case}}
            \put(217.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
            \put(257.1927710843373,35.0){\vector(0,-1){5.0}}
            \put(202.5,66.33333333333334){{\tiny obl}}
            \put(171.0,90.0){\vector(0,-1){60.0}}
            \put(176.0,80.0){{\tiny root}}
          \end{picture}
    \end{subfigure}
    \caption[The UD trees of an example active sentence and its passive counterpart]{The UD trees of an example active sentence and its passive counterpart.}
    \label{actpass}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}{.7\textwidth}
        \centering
        \includegraphics[width=.4\linewidth]{figure/different.pdf}
    \end{subfigure}%
    \begin{subfigure}{.3\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
        \begin{picture}(194.0,90.0)
          \put(0.0,0.0){they}
          \put(46.0,0.0){are}
          \put(83.0,0.0){different}
          \put(0.0,15.0){{\tiny PRON}}
          \put(46.0,15.0){{\tiny AUX}}
          \put(83.0,15.0){{\tiny ADJ}}
          \put(51.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
          \put(11.807228915662648,35.0){\vector(0,-1){5.0}}
          \put(36.5,66.33333333333334){{\tiny nsubj}}
          \put(74.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
          \put(60.054054054054056,35.0){\vector(0,-1){5.0}}
          \put(59.5,49.66666666666667){{\tiny cop}}
          \put(98.0,90.0){\vector(0,-1){60.0}}
          \put(103.0,80.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A GF AST and its UD counterpart]{A GF AST and its UD counterpart. Note how the copula is implicit in the GF AST and subordinate to its complement in UD.}
    \label{parsetree}
\end{figure}

@@

\subsection{Alignments} \label{aligns}
In terms of parallel text analysis, an \textit{alignment} consists in a pair of semantically equivalent (i.e., in this context, sharing the same abstract syntax) concrete expressions, one in the source and one in the target language. 
Even though, unless otherwise specified, we will keep referring to \textit{language pairs}, with a source and a target language, it is easy to see how this can be generalized to a more-than-bilingual, multidirectional definition: the pair just mentioned simply becomes a tuple of equivalent expressions in different languages. Formally, \smallskip

@@

\begin{itemize}
    \item Model 1 is based exclusively on lexical translation probabilities
    \item Model 2 takes word order (i.e. the absolute word positions) into account 
    \item Model 3 takes even a \textit{fertility} parameter into account (fertility being the number of target words that can be generated from a given source word)
    \item Models 4 and 5 takes the \textit{context} (i.e. the relative position of the words) in which each word occurs into account in a broader sense
\end{itemize}

@@

However, it must be noted that, in the context of statistical MT, the term ``phrase'' is not to be intended in its grammatical sense, but can refer to any sequence of - typically contiguous - words and in any case, both for word and phrase alignment, the relation that is established with these statistical methods is merely between strings in different languages: there is no intermediate representation capturing the concepts themselves. \smallskip

@@

Over time, various constituency grammar-based approaches have been proposed \cite{t2t1, t2t2}. These approaches, which make use of either parallel - manually or automatically annotated - treebanks, can be referred to as \textit{tree-to-tree alignment} methods \cite{bitext}. While theoretically appealing, however, their use in data-driven MT is limited, since they tend to suffer not only from the scarce availability of large-scale manually annotated corpora and the lack of robustness of the existing parsers, but also from the fact that, in general, the monolingual grammars used for parsing tend to be designed independently from each other, following different traditions, formalisms and linguistic theories \cite{nlphandbook}, thus making it often impossible to find a common representation describing a complete mapping from one tree to another \cite{bitext}.

\subsubsection{Our approach} \label{ourapproach}
Given that GF solves the latter problem, that it provides a good basis for NLG and that the intermediate representation it makes use of preserves all the syntactic information present in a sentence, the idea of trying to perform CA by comparing GF ASTs seems tempting. However, since ASTs need to be obtained from raw - or, at most, barely sentence-segmented - text, CA would suffer from the same problem that, as mentioned in in Section \ref{gf}, affects other tree-to-tree alignment methods and GF-based open-domain translation: the lack of a sufficiently robust analysis stage and, as a consequence, the inadequacy of the resulting parse trees. \smallskip

As anticipated throughout this chapter, we attempt to solve this problem by taking advantage of dependency parsing. In particular, UD is our formalism of choice since its focus on ``universality'' (i.e. on abstracting away from cross-lingual differences) makes it as interesting as GF itself when it comes to the possibility of establishing mappings between trees in different languages. Furthermore, when it comes to MT itself, using UD allows us to take advantage of the tools that have been developed with the intention of leveraging its similarities with GF (cf. Section \ref{ud}), thus enabling GF target language generation. \smallskip

@@

\begin{itemize}
    \item a UD parser
    \item an alignment module based on dependency tree comparison, which is the tangible output of this project and whose CE and CP components are described and evaluated in detail in Chapters \ref{ch3} and \ref{ch4} respectively
    \item a program, based on \texttt{gf-ud}, that converts the alignments produced in step 2 into GF ASTs to then generate a domain-specific GF lexicon\footnote{CA is performed on UD trees, i.e. before conversion to GF, to reduce the possibility of errors. The reason is that, as mentioned in Section \ref{udrelgf}, in fact, such conversion is not obtained by means of a deterministic procedure and, as such, is not always guaranteed to be correct.}
\end{itemize} \smallskip

>>>>

While many artificial languages, such as programming languages, are compositional by construction, the general validity of this principle for natural languages is, despite many arguments in favor, under debate. 
In this work, we make the assumption that compositionality also holds for natural languages, or at least that it is a useful notion applying to most natural language sentences\footnote{in Section \ref{evalign}, however, we will see that idiomatic expressions, often used as an arguments against compositionality, are also challenging to deal with in the context of grammar-based CA.}. \smallskip

@@

% intro to GF
Grammatical Framework (GF), being a grammar formalism and programming language designed with this parallel in mind, goes a step further in the direction of synchronous CFGs and TAGs. 
It introduces in fact a clear distinction between the \textit{abstract syntax}, whose aim is to capture the syntactic structures all natural languages taken into account have in common, and the \textit{concrete syntaxes}, specific to each individual language, consisting in their linearization rules \cite{gf2004} \cite{gfbook}. 
This makes it possible to deal with multiple languages by writing only one grammar, whose components are an abstract syntax and several concrete syntaxes, providing a solid basis for a system able to translate between any pair of for which a concrete syntax is available, in any direction. \smallskip

% GF as an interlingua-based MT system
With respects to this, GF can be also seen as an \textit{interlingual} MT system where the intermediate representation or \textit{interlingua} is an AST. Interlingua-based systems have the advantage, in terms of efficiency, of making it unnecessary to build $n (n - 1)$ translation functions to cover all possible pairs of $n$ languages: having $2n$ is in general enough. In GF in particular, using a multilingual grammar makes it so that translation can be performed, for each pair of languages, in both directions, so that $n$ translation functions (or rahter $n$ concrete syntaxes and an abstract syntax) are already sufficient.\smallskip

@@

\subsection{Dependency grammars} \label{depg}
% DGs as opposed to PSGs
A class of grammar formalisms alternative to phrase structure, first proposed in \cite{dg}, is that of \textit{dependency grammars}, the main difference between the two being the relation they are based on. 
As opposed to constituency, \textit{dependency} is a word-to-word correspondence, meaning that words are simply put in relation with each other via directed links, called in fact \textit{dependencies}. 
Each dependency is, then, composed of two words: a \textit{head} and a \textit{dependent} that refers to it. For instance, if we try to look at syntactic dependencies in the sentence ``\textit{grammars rule}'', we could identify the verb ``\textit{rule}'' as the head and the noun ``\textit{grammars}'', its subject, as its dependent. \smallskip

% pros
Intuitively, this means that dependency trees are simpler than their phrase structure counterparts. This makes them an easier target for the frontend, possibly ML-based, of a MT system such as the one outlined in the above. 
Existing \textit{dependency parsers}, such as UDPipe \cite{udpipe1} and the Standford parser \cite{standford}, are often - but not always \cite{rasp} - neural pipelines trained on dependency treebanks, significantly more robust than their phrase structure counterparts.

% cons
On the other hand, there is currently no effective way to use these trees as a starting point for NLG: dependency grammars are an effective way to describe language, but, unlike phrase-structure grammars, they are not \textit{generative},\footnote{that does not mean, however, that dependency grammars have not been made use of in MT research, in conjunction with statistical techniques, in order to better capture grammatical generalizations \cite{treelet}.}. \smallskip

One idea is, then, to develop a hybrid system where the frontend is a dependency parser and the backend a grammaticality-preserving GF-based target language generation module. 
This work goes in this direction, and the connecting link between these two seemingly incompatible stages of the pipeline is, as we will elaborate on in Section \ref{ourapproach}, a CA component able to identify matching dependency trees and generate the corresponding GF concrete and abstract syntax functions: concepts.

@@

In the following, we give a quick overview of the standard format UD uses to store dependency trees and of the aspects of the annotation scheme that are most relevant to the task at hand. 
Appendix \ref{a} provides a more comprehensive, but not completely exhaustive, description of all the POS tags and dependency labels mentioned in the examples appearing in this work based on that in \cite{compsyn}. 
The reader interested in a full specification of the UD annotation scheme may refer to the official UD documentation\footnote{available at \url{universaldependencies.org}.}.

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.55\textwidth}
      \centering
      \footnotesize
        \begin{verbatim}
1	this	this	PRON	_	_	5	nsubj	_	_
2	is	be	AUX	_	_	5	cop	_	_
3	a	a	DET	_	_	5	det	_	_
4	dependency	dependency	NOUN	_	_	5	compound	_	_
5	tree	tree	NOUN	_	_	0	root	_	_
        \end{verbatim}
    \end{subfigure}%
    \begin{subfigure}{.45\textwidth}
        \centering
        \footnotesize
        \setlength{\unitlength}{0.21mm}
        \begin{picture}(371.0,130.0)
            \put(0.0,0.0){this}
            \put(46.0,0.0){is}
            \put(83.0,0.0){a}
            \put(120.0,0.0){dependency}
            \put(220.0,0.0){tree}
            \put(0.0,15.0){{\tiny PRON}}
            \put(46.0,15.0){{\tiny AUX}}
            \put(83.0,15.0){{\tiny DET}}
            \put(120.0,15.0){{\tiny NOUN}}
            \put(220.0,15.0){{\tiny NOUN}}
            \put(120.0,30.0){\oval(218.63636363636363,133.33333333333334)[t]}
            \put(10.681818181818187,35.0){\vector(0,-1){5.0}}
            \put(105.0,99.66666666666667){{\tiny nsubj}}
            \put(143.0,30.0){\oval(172.27586206896552,100.0)[t]}
            \put(56.86206896551724,35.0){\vector(0,-1){5.0}}
            \put(128.0,83.0){{\tiny cop}}
            \put(161.5,30.0){\oval(134.8102189781022,66.66666666666667)[t]}
            \put(94.0948905109489,35.0){\vector(0,-1){5.0}}
            \put(146.5,66.33333333333334){{\tiny det}}
            \put(180.0,30.0){\oval(97.0,33.333333333333336)[t]}
            \put(131.5,35.0){\vector(0,-1){5.0}}
            \put(165.0,49.66666666666667){{\tiny compound}}
            \put(235.0,130.0){\vector(0,-1){100.0}}
            \put(240.0,120.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A dependency tree in CoNNL-U format alongside its graphical representation]{A dependency tree in CoNNL-U format alongside its graphical representation. Optional fields in the CoNLL-U file are left blank.}
    \label{deptree}
\end{figure}

\paragraph{Universal POS tags} \label{upos}
Universal POS tags mark the core part-of-speech categories, such as nouns, verbs, pronouns and determiners. \smallskip

An important distinction we can make based on POS tags is, as we will see in Section \ref{criteria}, that between \textit{content} and \textit{function} words.
\textit{Content} or \textit{open class} words are words with a lexical meaning (nouns, lexical verbs, adjectives, adverbs and interjections).
\textit{Function} words, like pronouns and determiners, only have a grammatical meaning. Since they do not readily accept new members\footnote{one exception is the recent introduction of the gender-neutral pronoun ``\textit{hen}'' in Swedish.}, they are often referred to also as \textit{closed class} words. \smallskip

@@

 If a sentence contains is no verbs at all, there is no fixed rule excepts that, in order to avoid unnecessary discrepancies between languages, the root should be a content word. \smallskip

\begin{figure}[h]

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
\begin{picture}(259.0,90.0)
  \put(0.0,0.0){Katia}
  \put(55.0,0.0){reads}
  \put(110.0,0.0){a}
  \put(147.0,0.0){magazine}
  \put(0.0,15.0){{\tiny PROPN}}
  \put(55.0,15.0){{\tiny VERB}}
  \put(110.0,15.0){{\tiny DET}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(37.5,30.0){\oval(49.54545454545455,33.333333333333336)[t]}
  \put(12.727272727272727,35.0){\vector(0,-1){5.0}}
  \put(22.5,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny det}}
  \put(121.0,30.0){\oval(88.73913043478261,66.66666666666667)[t]}
  \put(165.3695652173913,35.0){\vector(0,-1){5.0}}
  \put(106.0,66.33333333333334){{\tiny obj}}
  \put(70.0,90.0){\vector(0,-1){60.0}}
  \put(75.0,80.0){{\tiny root}}
\end{picture}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
        \begin{picture}(344.0,90.0)
            \put(0.0,0.0){A}
            \put(37.0,0.0){magazine}
            \put(119.0,0.0){is}
            \put(156.0,0.0){read}
            \put(202.0,0.0){by}
            \put(239.0,0.0){Katia}
            \put(0.0,15.0){{\tiny DET}}
            \put(37.0,15.0){{\tiny NOUN}}
            \put(119.0,15.0){{\tiny AUX}}
            \put(156.0,15.0){{\tiny VERB}}
            \put(202.0,15.0){{\tiny ADP}}
            \put(239.0,15.0){{\tiny PROPN}}
            \put(28.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(14.054054054054054,35.0){\vector(0,-1){5.0}}
            \put(13.5,49.66666666666667){{\tiny det}}
            \put(106.5,30.0){\oval(116.47899159663865,66.66666666666667)[t]}
            \put(48.260504201680675,35.0){\vector(0,-1){5.0}}
            \put(91.5,66.33333333333334){{\tiny nsubj:pass}}
            \put(147.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(133.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(132.5,49.66666666666667){{\tiny aux:pass}}
            \put(230.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
            \put(216.05405405405406,35.0){\vector(0,-1){5.0}}
            \put(215.5,49.66666666666667){{\tiny case}}
            \put(217.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
            \put(257.1927710843373,35.0){\vector(0,-1){5.0}}
            \put(202.5,66.33333333333334){{\tiny obl}}
            \put(171.0,90.0){\vector(0,-1){60.0}}
            \put(176.0,80.0){{\tiny root}}
          \end{picture}
    \end{subfigure}
    \caption[The UD trees of an example active sentence and its passive counterpart]{The UD trees of an example active sentence and its passive counterpart.}
    \label{actpass}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.7\textwidth}
        \centering
        \includegraphics[width=.4\linewidth]{figure/different.pdf}
    \end{subfigure}%
    \begin{subfigure}{.3\textwidth}
        \centering
        \setlength{\unitlength}{0.25mm}
        \begin{picture}(194.0,90.0)
          \put(0.0,0.0){they}
          \put(46.0,0.0){are}
          \put(83.0,0.0){different}
          \put(0.0,15.0){{\tiny PRON}}
          \put(46.0,15.0){{\tiny AUX}}
          \put(83.0,15.0){{\tiny ADJ}}
          \put(51.5,30.0){\oval(79.3855421686747,66.66666666666667)[t]}
          \put(11.807228915662648,35.0){\vector(0,-1){5.0}}
          \put(36.5,66.33333333333334){{\tiny nsubj}}
          \put(74.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
          \put(60.054054054054056,35.0){\vector(0,-1){5.0}}
          \put(59.5,49.66666666666667){{\tiny cop}}
          \put(98.0,90.0){\vector(0,-1){60.0}}
          \put(103.0,80.0){{\tiny root}}
        \end{picture}
    \end{subfigure}
    \caption[A GF AST and its UD counterpart]{A GF AST and its UD counterpart. Note how the copula is implicit in the GF AST and subordinate to its complement in UD.}
    \label{parsetree}
\end{figure}

@@

\subsection{Alignments} \label{aligns}
In terms of parallel text analysis, an \textit{alignment} consists in a pair of semantically equivalent (i.e., in this context, sharing the same abstract syntax) concrete expressions, one in the source and one in the target language. 
Even though, unless otherwise specified, we will keep referring to \textit{language pairs}, with a source and a target language, it is easy to see how this can be generalized to a more-than-bilingual, multidirectional definition.
The pair just mentioned simply becomes a tuple of equivalent expressions in different languages. Formally, \smallskip

@@

\begin{itemize}
    \item Model 1 is based exclusively on lexical translation probabilities
    \item Model 2 takes word order (i.e. the absolute word positions) into account 
    \item Model 3 takes even a \textit{fertility} parameter into account (fertility being the number of target words that can be generated from a given source word)
    \item Models 4 and 5 take the \textit{context} (i.e. the relative position of the words) in which each word occurs into account in a broader sense
\end{itemize}

@@

However, it must be noted that, in the context of statistical MT, the term ``phrase'' is not to be intended in its grammatical sense, but can refer to any sequence of - typically contiguous - words. 
In any case, both for word and phrase alignment, the relation that is established with these statistical methods is merely between strings in different languages: there is no intermediate representation capturing the concepts themselves. \smallskip

@@

Over time, various constituency grammar-based approaches have been proposed \cite{t2t1, t2t2}. 
These approaches, which make use of parallel treebanks, can be referred to as \textit{tree-to-tree alignment} methods \cite{bitext}. 
While theoretically appealing, however, their use in data-driven MT is limited, since they tend to suffer not only from the scarce availability of large-scale manually annotated corpora and the lack of robustness of the existing parsers, but also from the fact that, in general, the monolingual grammars used for parsing tend to be designed independently from each other, following different traditions, formalisms and linguistic theories \cite{nlphandbook}, thus making it often impossible to find a common representation describing a complete mapping from one tree to another \cite{bitext}.

\subsubsection{Our approach} \label{ourapproach}
Given that GF solves the latter problem, that it provides a good basis for NLG and that the intermediate representations it makes use of preserve all the syntactic information present in a sentence, the idea of trying to perform CA by comparing GF ASTs seems tempting. 
However, since ASTs need to be obtained from raw - or, at most, barely sentence-segmented - text, CA would suffer from the same problem that, as mentioned in in Section \ref{gf}, affects other tree-to-tree alignment methods and GF-based open-domain translation: the lack of a sufficiently robust analysis stage and, as a consequence, the inadequacy of the resulting parse trees. \smallskip

As anticipated throughout this chapter, we attempt to solve this problem by taking advantage of dependency parsing. 
In particular, UD is our formalism of choice since its focus on ``universality'' (i.e. on abstracting away from cross-lingual differences) makes it as interesting as GF itself when it comes to the possibility of establishing mappings between trees in different languages. 
Furthermore, when it comes to MT itself, using UD allows us to take advantage of the tools that have been developed to leverage its similarities with GF (cf. Section \ref{ud}), thus enabling GF target language generation. \smallskip

@@

\begin{itemize}
    \item a UD parser
    \item an alignment module based on dependency tree comparison, which is the tangible output of this project and whose CE and CP components are described and evaluated in detail in Chapters \ref{ch3} and \ref{ch4} respectively
    \item a program, based on \texttt{gf-ud}, that converts the alignments into GF ASTs to then generate a domain-specific GF lexicon\footnote{CA is performed on UD trees, i.e. before conversion to GF, to reduce the possibility of errors. The reason is that, as mentioned in Section \ref{udrelgf}, in fact, such conversion is not obtained by means of a deterministic procedure and, as such, is not always guaranteed to be correct.}.
\end{itemize} \smallskip
